---
title: "Phasing"
output:
pdf_document:
  toc: false
  highlight: zenburn
---

```{r}
load_all("~/Projects/ProgenyArray")
library(rmarkdown)
library(Rcpp)
library(parallel)
library(caret)
library(reshape2)
options(mc.cores=4)
set.seed(42)
```

## Simulate Sibling Family with Selfs

```{r}
sf <- sibFamily(10, 1000)
```

How do these separate in principal component space?

```{r}
p <- prcomp(t(sf$progeny))
d <- data.frame(x=p$x[,1], y=p$x[,2], mom=as.factor(sf$metadata$mgamete),
                type=sf$metadata$type)
ggplot(d) + geom_point(aes(x=x, y=y, color=mom, shape=type)) + xlab("PC 1") + ylab("PC 2")

```

## K-means

Let's use k-means to look at accuracy. K-means is non-deterministic (random
initial positions), so we will run 1000 replicates. Cluster labels are
arbitrary, so we swap them when the alternate label is closer.

```{r}

switchLabels <- function(inferred, real) {
  switch <- which.max(table(inferred, real)[1, ]) != 1
  if (switch)
    return(ifelse(inferred == 1, 2, 1))
  inferred
}

kmeansAccuracy <- function(geno, real) {
  cm <- confusionMatrix(switchLabels(kmeans(geno, 2)$cluster, real), real)
  cm$overall['Accuracy']
}

famsize <- rep(5:70, each=3)
acc <- sapply(famsize, function(n) {
                 sf <- sibFamily(n, 1000)
                 kmeansAccuracy(t(sf$progeny), sf$metadata$mgamete)
})

qplot(x=famsize, y=acc) + ylab("accuracy") + xlab("family size") + geom_smooth(se=FALSE)

```

What is being misclassified? These are all heterozygous selfed progeny, where
the true cluster is either.

```{r}
incor <- sapply(famsize, function(n) {
                sf <- sibFamily(n, 1000)
                l <- switchLabels(kmeans(t(sf$progeny), 2)$cluster,
                                  sf$metadata$mgamete)
                nwrong <- sum(l != sf$metadata$mgamete)
                # how many cases are wrong, but heterzygous selfs
                x <- with(sf$metadata, l != mgamete & type == "selfed"
                          & mother != father)
                sum(x)/nwrong
})

```

## Imputation

```{r}

haploImputeWindow <- function(geno, parent_geno, fathers, freqs, fun=kmeans, ehet=0.8, ehom=0.1) {
  geno_na <- na.omit(geno)
  if (nrow(geno_na) < 1) {
    warning("cannot run k-means on this window: all loci contian NA values")
    return(NULL)
  }
  nremoved <- length(na.action(geno_na))
  #if (nremoved > 1) warning(sprintf("%d ind")
  km <- fun(t(geno_na), 2)
  hap_geno <- list(geno[, km$cluster == 1], geno[, km$cluster == 2])
  out <- lapply(1:2, function(which_hap) {
                # iterate haplotype imputation for each cluster
                hap_geno <- geno[, km$cluster == which_hap, drop=FALSE]
                hap_fathers <- fathers[km$cluster == which_hap]
                ll <- lapply(1:nrow(hap_geno), function(i) {
                             allele_ll(hap_geno[i, ], ehet, ehom, freqs[i],
                                       parent_geno[cbind(i, hap_fathers)])
               })
               do.call(rbind, ll)
  })
}





# example run:

ll <- haploImputeWindow(sf$progeny, sf$geno, sf$metadata$father, sf$freqs, ehet=0.1, ehom=0.1)

mom <- sf$haplos[[1]]
h1 <- apply(ll[[1]], 1, which.max) - 1L
h2 <- apply(ll[[2]], 1, which.max) - 1L

```

Variable family size.

```{r}
famsizes <- c(5, seq(10, 80, 10))

hamming <- function(x, y) sum(x != y)

hapConfusion <- function(inferred, real) {
  if (hamming(inferred[, 1], real[, 1]) < hamming(inferred[, 1], real[, 2])) {
    out <- list(confusionMatrix(inferred[, 1], real[, 1]),
                confusionMatrix(inferred[, 2], real[, 2]))
  } else {
    out <- list(confusionMatrix(inferred[, 1], real[, 2]),
                confusionMatrix(inferred[, 2], real[, 1]))
  }
  return(out)
}

accs <- lapply(famsizes, function(fs) {
 replicate(10, {
           sf <- sibFamily(fs, 1000)
           ll <- haploImputeWindow(sf$progeny, sf$geno, sf$metadata$father, sf$freqs, ehet=0.1, ehom=0.1)
           mom <- sf$haplos[[1]]
           h1 <- apply(ll[[1]], 1, which.max) - 1L
           h2 <- apply(ll[[2]], 1, which.max) - 1L
           cm <- hapConfusion(cbind(h1, h2), mom)
           sapply(cm, function(x) x$overall['Accuracy'])
  })
})

# each list element is a matrix, with two rows, each row both haplotypes'
# accuracies. So we munge:
reshapeReplicatesSimResults <- function(results, var, varname) {
  tmp <- do.call(rbind, results)
  d <- as.data.frame(tmp, row.names=1:nrow(tmp))
  d[, varname] <- rep(var, each=2)
  melt(d, id=varname)[, c(varname, 'value')]
}

d <- reshapeReplicatesSimResults(accs, famsizes, "famsize")
p <- ggplot(d, aes(x=famsize, y=value)) + geom_point() + geom_smooth(se=FALSE)
p <- p + xlab("family size") + ylab("imputation accuracy")
print(p)

```

or using boxplot:

```{r}
p <- ggplot(d) + geom_boxplot(aes(x=as.factor(famsize), y=value))
p <- p + xlab("family size") + ylab("imputation accuracy")
p <- p + scale_y_continuous(limits=c(0, 1))
print(p)

```

## With cclust's cclust

From the `cclust` package.


```{r}
famsizes <- c(5, seq(10, 80, 10))

hapConfusion <- function(inferred, real) {
  if (hamming(inferred[, 1], real[, 1]) < hamming(inferred[, 1], real[, 2])) {
    out <- list(confusionMatrix(inferred[, 1], real[, 1]),
                confusionMatrix(inferred[, 2], real[, 2]))
  } else {
    out <- list(confusionMatrix(inferred[, 1], real[, 2]),
                confusionMatrix(inferred[, 2], real[, 1]))
  }
  return(out)
}

accs <- lapply(famsizes, function(fs) {
 replicate(10, {
           sf <- sibFamily(fs, 1000)
           llkm <- haploImputeWindow(sf$progeny, sf$geno, sf$metadata$father, sf$freqs, ehet=0.1, ehom=0.1)
           llcc <- haploImputeWindow(sf$progeny, sf$geno, sf$metadata$father, sf$freqs, ehet=0.1, ehom=0.1)
           mom <- sf$haplos[[1]]
           h1km <- apply(llkm[[1]], 1, which.max) - 1L
           h2km <- apply(llkm[[2]], 1, which.max) - 1L
           h1cc <- apply(llcc[[1]], 1, which.max) - 1L
           h2cc <- apply(llcc[[2]], 1, which.max) - 1L
           cmkm <- hapConfusion(cbind(h1km, h2km), mom)
           cmcc <- hapConfusion(cbind(h1cc, h2cc), mom)
           setNames(sapply(c(cmkm, cmcc), function(x) x$overall['Accuracy']), c("km", "km", "cc", "cc"))
  })
})

# munge data together
tmp <- do.call(rbind, accs)
d <- as.data.frame(tmp, row.names=1:nrow(tmp))
d$method <- rep(c("kmeans", "kmeans", "cclust", "cclust"), rep=nrow(tmp)/4) # FIX
d$famsize <- rep(famsizes, each=4)
dm <- melt(d, id=c("method", "famsize"))[, c("famsize", "method", "value")]
ggplot(dm, aes(x=famsize, y=value, color=method)) + geom_point() + geom_smooth(se=FALSE)

}
```

Therefore, not much of a difference - we use cclust because it's faster and has
better prediction abilities:

```
# on large sib family 100,000 loci
system.time(kmeans(t(sf$genos), 2))
 user  system elapsed
 0.189   0.012   0.201
system.time(cclust(t(sf$genos), 2))
 user  system elapsed
 0.041   0.003   0.043
```

## With Missing Data

```{r}

missing <- seq(0.01, 0.15, by=0.01)
na_accs <- lapply(missing, function(m) {
 replicate(10, {
           sf <- sibFamily(30, 1000, na_rate=m)
           ll <- haploImputeWindow(sf$progeny, sf$geno, sf$metadata$father, sf$freqs, ehet=0.1, ehom=0.1)
           mom <- sf$haplos[[1]]
           h1 <- apply(ll[[1]], 1, which.max) - 1L
           h2 <- apply(ll[[2]], 1, which.max) - 1L
           cm <- hapConfusion(cbind(h1, h2), mom)
           sapply(cm, function(x) x$overall['Accuracy'])
  })
})

# each list element is a matrix, with two rows, each row both haplotypes'
# accuracies. So we munge:

d <- reshapeReplicatesSimResults(na_accs, missing, "missing")
p <- ggplot(d, aes(x=missing, y=value)) + geom_point() + geom_smooth(se=FALSE, color="red")
p <- p + xlab("proportion missing at random") + ylab("imputation accuracy") + ggtitle("haplotype imputation accuracy (famsize=30)")
print(p)

```
